The Poisson approximate likelihood compared to the particle filter

### 2024-08-22 ms.Rnw
comment on initial conditions, presumably delete Figure~2 later

\cite{stocks} fitted their model to a dataset derived by dividing the original reported count data by an estimated reporting rate, to put their data on the scale of the actual number of cases in the population, whereas \cite{wwr} fitted directly to the report data.
The reporting rate used by \cite{stocks} varied over time, but was generally around $7\%$.
On approximately 1200 data points, this corresponds to a discrepancy of around $-1200\log(0.07) \approx 3200$ log-likelihood units, largely explaining the difference reported in Table~1 and interpreted by \cite{wwr} as evidence supporting PAL.
The comparison can be corrected either by applying the method of SBH to the data of WWR or vice versa.
Since the method of SBH is applicable to a more general class of models, and supported by published software, it was convenient to apply the SBH method to the model and data of WWR.
The large discrepancy in log-likelihood disappears at this point by recomputing the likelihood of the model using PAL and particle filter separately (see Table~\ref{tab:ovovrealdata}).
This re-analysis does, however, show a discrepancy between two methods.
We continued our investigation to establish the cause of this.
We discovered that WWR (unlike SBH) used a poorly fitting choice of initial conditions for the latent process in January 2001 (their time 0).

>>>>

\cite{stocks} fitted their model to a dataset derived by dividing the original reported count data by an estimated reporting rate, to put their data on the scale of the actual number of cases in the population, whereas \cite{wwr} fitted directly to the report data.
The reporting rate used by \cite{stocks} varied over time, but was generally around $7\%$.
On approximately 1200 data points, this corresponds to a discrepancy of around $-1200\log(0.07) \approx 3200$ log-likelihood units, largely explaining the difference reported in Table~1 and interpreted by \cite{wwr} as evidence supporting PAL.
The comparison can be corrected either by applying the method of SBH to the data of WWR or vice versa.
Since the method of SBH is applicable to a more general class of models, and supported by published software, it was convenient to apply the SBH method to the model and data of WWR.
The large discrepancy in log-likelihood disappears at this point by recomputing the likelihood of the model using PAL and particle filter separately (see Table~\ref{tab:ovovrealdata}).
This re-analysis does, however, show a discrepancy between two methods.
We continued our investigation to establish the cause of this.
We discovered that the initial conditions for the latent process in January 2001 (their time 0) were estimated by SBH. However, in WWR, they were assumed fixed and made the model fitted poorly.
